---
title: "Figures for Herp. Cons. Bio Manuscript - Code extracted from Kauphusman Thesis"
author: "C.M. Gienger"
date: "6/10/2026"
output:
    rmdformats::readthedown:
      self_contained: true
      thumbnails: true
      lightbox: true
      gallery: false
      highlight: tango
      use_bookdown: true
      fig_height: 12
      fig_width: 12
      html_document:
      code_folding: "hide"
      
---

```{r Library, include=FALSE}
packages <- c(
  "caret",
  "utils",
  "raster",
  "data.table",
  "grid",
  "vegan",
  "permute",
  "sp",
  "dplyr",
  "ggplot2",
  "tidyr",
  "knitr",
  "lattice",
  "MASS",
  "randomForest",
  "randomForestExplainer",
  "ROCR",
  "cowplot",
  "ggrepel",
  "gt",
  "glue",
  "pdp",
  "scales",
  "kableExtra",
  "cowplot",
  "vegan"
)

sapply(packages, require, character.only=T)
```

```{r set font for Herp Cons Bio}
theme_set(
  theme_bw(base_family = "Times New Roman")
)
```

```{r Dataset, message=FALSE, warning=FALSE}
study <- read.csv("public_data/Cottonmouth-Datasheet.csv")
```

### Hydrology Plots

```{r Hydro Graphs, message=FALSE, warning=FALSE}
gauge<-read.csv("data/GaugeData-USGS.csv")
gauge[,2]<-as.Date(gauge[,2])
gauge[,4]<-gauge[,3]*0.3048

discharge<-read.csv("data/SycamoreData-USGS.csv")
discharge[,2]<-as.Date(discharge[,2])
discharge[,4]<-discharge[,3]*0.0283168

p1<-ggplot(gauge, aes(x =gauge$datetime, y= gauge$V4))+
  geom_line(color = "black")+
  xlab("")+
  ylab("Gauge Height (m)")+
  ggtitle("Cumberland River Below Cheatham Dam, TN")+
  #ggtitle("USGS 03435000 Cumberland River Below Cheatham Dam, TN")+
  scale_x_date(breaks=date_breaks("6 months"), labels = date_format("%b %y"))+
  theme_classic() +
  theme(text = element_text(family = "Times New Roman")) +
  theme(
    plot.title = element_text(size = 14),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(size = 12, color = "black"),
    axis.text.y = element_text(size = 12, color = "black")
  )

p2<-ggplot(discharge, aes(x =discharge$datetime, y= discharge$V4))+
  geom_line(color = "black")+
  xlab("")+
  ylab(expression("Discharge (" * m^3 * "/second)"))+
  ggtitle("Sycamore Creek Near Ashland City, TN")+
  #ggtitle("USGS 03431800 Sycamore Creek Near Ashland City, TN")+
  scale_x_date(breaks=date_breaks("6 months"), labels = date_format("%b %y"))+
  theme_classic() +
  theme(text = element_text(family = "Times New Roman")) +
  theme(
    plot.title = element_text(size = 14),
    axis.title.x = element_text(size = 12), 
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(size = 12, color = "black"),
    axis.text.y = element_text(size = 12, color = "black")
  )

(p1 / p2) #stack plots using patchwork

ggsave("HCB_figs/fig2_hydrographs.jpg")
#ggsave("HCB_figs/fig1_hydrographs.jpg", width = 3.43, height = 3)
```

**Figure 1** Hydrographs of the annual hydrology of the Cumberland River below Cheatham Dam, TN recorded at 36°19'22.26" N & 87°13'41.73" W, and Sycamore Creek Near Ashland City, TN (below) recorded at 36°19'11.89" N, & 87°03'04.29"W. Both waterways flow into Cumberland River Bicentennial Trail riverine wetland slough and were recorded within 8 km from the study site. Information was provided by the USGS National Water Information System (https://waterdata.usgs.gov/nwis).

# **Results**
```{r General Statistics, message=FALSE, warning=FALSE}
#Subsetting Groups
male<- study %>% subset(Group == "MALE")
female<- study %>% subset(Group == "FEMALE")
juvenile<- study %>% subset(Group == "JUVENILE")
YOY<- study %>% subset(Group == "YOY")
```

## Habitat Selection: Ecological Niche 

List of the variables of the dataset for Multivariate and random forest analyses. 

```{r Variable Exclusion, message=FALSE, warning=FALSE, include=FALSE}
glimpse(study)
s<-study[,c(3:24)]
s<-na.omit(s)
glimpse(s)
```

```{r Drop Column Feature Selection, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
#Correlation Check
corrrel<-cor(s[,c(2:20)])
findCorrelation(corrrel, cutoff = 0.7)
```

```{r Removing correlated features, include=FALSE}
#Averaged surface and ambient temperature variables
Temp<-(s$Surface.Temp+s$Ambient.Temp)/2
s<-cbind(s,Temp)
#Removed correlated variables
s<-s[,c(1,5,6,7,8,9,10,11,12,13,14,15,17,18,19,20,21,22,23)]
glimpse(s)
```

```{r Multivariate Analysis, fig.height=9, fig.width=10, message=FALSE, warning=FALSE}
s1<-s[,c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19)]
glimpse(s1)
#PCA
pc<-princomp(s1[,c(2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,19)], cor = T, scores = T)
#Component
#Setting- Up ANOVA
comp<-pc$scores[,1:3]
s1<-cbind(s1, comp)
#Snake Averages
group<- s1 %>% subset(Type == "SNAKE")

pcomp<-group_by(group, Group)%>%
  summarise(Comp.1 = mean(Comp.1),
            Comp.2 = mean(Comp.2),
            Comp.3 = mean(Comp.3))
```
The PCA summary table with scree methods to determine components retained.

```{r Scree Method, fig.height=4, fig.width=4, message=FALSE, warning=FALSE}
summary(pc)
screeplot(pc, type = "l")
```

The scree method retains eigenvalues that explain more variance than would be expected by random. Therefore, the largest eigenvalues above the average eigenvalue score, along with the first subsequent component below the largest eigenvalues, are retained (Jackson, 1993). 

### PCA

```{r Multivariate PCA Plots, fig.height=8, fig.width=12, message=FALSE, warning=FALSE}
##Colored Plot Snake vs Habitat
p<-ggplot(data = s1, aes(Comp.1, Comp.2, shape = Type))+
  geom_point(size = 3.1)+
  scale_shape_manual(values = c(21, 19))+
  theme_bw(base_size = 24)+
  theme(panel.grid = element_blank(), panel.border = element_rect(fill= "transparent"))+
  geom_hline(yintercept = 0, lty = 2) +
  geom_vline(xintercept = 0, lty = 2)+
  xlim(-6,6)+
  ylim(-4.5,4.5)+
  stat_ellipse(type = "norm", aes(color = Type), 
               show.legend = FALSE,size = 1, level = 0.95, linetype = 1)+
  scale_color_manual(values=c("#969696", "#252525"))+
  annotate("label",  x = 3.5, y = 2.7, label = "Cottonmouth", color ="#252525", size = 9, family = "Times New Roman")+
  annotate("label",  x = 3, y = -3.5, label = "Habitat", color ="#969696", size = 9, family = "Times New Roman")+
  xlab(expression(atop("aquatic sites" %<->% "terrestrial sites", paste("PC-1 (22.6%)" ))))+
  ylab(expression(atop ("PC-2 (11.8%)", paste("low cover " %<->% "high cover"))))+
  theme(text = element_text(family = "Times New Roman")) +
  theme(legend.position="none", legend.background = element_rect(fill="White",
                                                                 size=0.5),
    axis.title.x = element_text(size = 26),
    axis.title.y = element_text(size = 26),
    axis.text.x = element_text(size = 20, color = "black"),
    axis.text.y = element_text(size = 20, color = "black")
  )

#Adding Densities
xdens<-axis_canvas(p, axis = "x")+
  geom_density(data =group, aes(Comp.1, color = group$Group, fill = group$Group), alpha = 0.15, size =1, linetype= "dashed")+
  scale_fill_manual(values = c("#F8766D","#7CAE00","#00BFC4", "#C77CFF"))+
  scale_color_manual(values = c("#F8766D","#7CAE00","#00BFC4", "#C77CFF")) +
  theme(text = element_text(family = "Times New Roman"))

ydens<-axis_canvas(p, axis = "y", coord_flip = TRUE)+
  geom_density(data =group, aes(Comp.2, color = group$Group, fill = group$Group), alpha = 0.15, size =1, linetype= "dashed")+
  scale_fill_manual(values = c("#F8766D","#7CAE00","#00BFC4", "#C77CFF"))+
  scale_color_manual(values = c("#F8766D","#7CAE00","#00BFC4", "#C77CFF"))+
  theme(text = element_text(family = "Times New Roman")) +
  coord_flip()

p1<-insert_xaxis_grob(p, xdens, grid::unit(0.2, "null"), position= "top")
p2<-insert_yaxis_grob(p1,ydens, grid::unit(0.2, "null"), position = "right")

twogroup<-ggdraw(p2)

#Colored Intraspecific Group
gg_scatter1<-ggplot(data = s1, aes(Comp.1, Comp.2, shape = Type))+
  scale_shape_manual(values = c(15,21,16,17,18, 19))+
  theme_bw(base_size = 24)+
  theme(text = element_text(family = "Times New Roman")) +
  theme(panel.grid = element_blank(), panel.border = element_rect(fill= "transparent"))+
  geom_hline(yintercept = 0, lty = 2) +
  geom_vline(xintercept = 0, lty = 2)+
  xlim(-6,6)+
  ylim(-4.5,4.5)+
  stat_ellipse(type = "norm", aes(color = Type), 
               show.legend = FALSE,size = 1, level = 0.95, linetype = 1)+
  scale_color_manual(values=c("#F8766D","#969696","#7CAE00","#00BFC4", "#C77CFF", "#252525"))+
  annotate("label",  x = 3.2, y = 2.7, label = "Cottonmouth", color ="#252525", size = 9, family = "Times New Roman")+
  annotate("label",  x = 3, y = -3.5, label = "Habitat", color ="#969696", size = 9, family = "Times New Roman")+
  xlab(expression(atop("aquatic sites" %<->% "terrestrial sites", paste("PC-1 (22.6%)" ))))+
  ylab(expression(atop ("PC-2 (11.8%)", paste("low cover " %<->% "high cover"))))+
  theme(legend.position="none", legend.background = element_rect(fill="White",
                                                                 size=0.5),
    axis.title.x = element_text(size = 26),
    axis.title.y = element_text(size = 26),
    axis.text.x = element_text(size = 20, color = "black"),
    axis.text.y = element_text(size = 20, color = "black"))

pp<-gg_scatter1+
  geom_point(data =pcomp, aes(pcomp$Comp.1, pcomp$Comp.2, shape = Group, color = Group), size = 8, alpha= 0.8)+
  theme(text = element_text(family = "Times New Roman")) +
  annotate("label", x = 0.9, y = 0.5, label = "FEMALE", color ="#F8766D", size = 6, family = "Times New Roman")+
  annotate("label", x = -1.6, y = 1.2, label = "JUVENILE", color ="#7CAE00", size = 6, family = "Times New Roman")+
  annotate("label", x = -0.7, y = -0.15, label = "MALE", color ="#00BFC4", size = 6, family = "Times New Roman")+
  annotate("label", x = 0.8, y = 1.7, label = "NEONATE", color ="#C77CFF", size =6, family = "Times New Roman")

#Adding Densities for Intraspecific Groups
xdens1<-axis_canvas(pp, axis = "x")+
  geom_density(data =group, aes(Comp.1, color = group$Group, fill = group$Group), alpha = 0.15, size =1, linetype= "dashed")+
  scale_fill_manual(values = c("#F8766D","#7CAE00","#00BFC4", "#C77CFF"))+
  scale_color_manual(values = c("#F8766D","#7CAE00","#00BFC4", "#C77CFF"))

ydens1<-axis_canvas(pp, axis = "y", coord_flip = TRUE)+
  geom_density(data =group, aes(Comp.2, color = group$Group, fill = group$Group), alpha = 0.15, size =1, linetype= "dashed")+
  scale_fill_manual(values = c("#F8766D","#7CAE00","#00BFC4", "#C77CFF"))+
  scale_color_manual(values = c("#F8766D","#7CAE00","#00BFC4", "#C77CFF"))+
  coord_flip()

p11<-insert_xaxis_grob(pp, xdens1, grid::unit(0.2, "null"), position= "top")
p22<-insert_yaxis_grob(p11,ydens1, grid::unit(0.2, "null"), position = "right")

fourgroup<-ggdraw(p22)

colored<-plot_grid(twogroup,fourgroup, nrow = 1, align = "h")
```

#### Colored PCA Plot
```{r PCA colored, fig.height=10, fig.width=22, message=FALSE, warning=FALSE}
colored
ggsave("HCB_figs/fig3_PCA_ellipses.jpg")
```

**Figure 2:** Principal Component Analysis (PCA) of 16 environmental variables measured for A. piscivorus (cottonmouth) and available habitat comparison sites. Ellipses represent 95% confidence intervals for cottonmouth (small grey points) and available habitat (small black points). Mean component scores of the intra-specific cottonmouth conspecifics; adult males (blue/diamond/dark grey), adult females (red/ large circle/ black), juveniles (green/ large cirlce/ light grey), and neonates (purple/ triangle/ medium grey). Point density curves for conspecific groups are represented for PC-1 (above), and PC-2 (right). Linetypes of the density curves are; neonates with dotted lines, juveniles with solid lines, males with dashed, and females with no outline. PCA expresses cottonmouths to share broad similarities with the available habitat comparison sites, while neonates expressed some degree of separation toward the other cottonmouth conspecifics.

## PCA Loadings and PERMANOVA/ PERMDISP

**Table 1:** Factor loadings from principal component analysis of environmental variables measured for cottonmouths and general habitat comparitsion sites

```{r Loadings Table, message=FALSE, warning=FALSE}
Description<- c("Percentage of canopy above quadrat", "Coverage within quadrat","Coverage within quadrat","Coverage within quadrat","Coverage within quadrat", "Height of largest living woody stem", "Height of cover object closest to snake point", "Distance to nearest cover object", "Diameter of largest log within quadrat", "Distance to nearest water source", "Water depth measured within quadrat", "Coverage within quadrat","Coverage within quadrat","Coverage within quadrat", "Coverage within quadrat","Mean temperature within quadrat", "", "")

loadings<-pc$loadings[,1:3]
sdev<-t(data.frame(pc$sdev))
eigen<-sdev[,1:3]
Variance<-c(22.60, 11.81, 9.27)
loadings<-rbind(loadings, eigen, Variance)
loadings<-round(loadings, digits = 3)
names(loadings)<-c("PC-1", "PC-2", "PC-3")
loadings<-round(loadings, digits = 3)
loadings<-cbind(Description, loadings)

"Variable (units)"<-c("Canopy Closure (%)", "Rock Cover (%)", "Leaf Cover (%)", "Vegetative Cover (%)", "Fallen Log Cover (%)","Woody Stem Height (cm)", "Cover Height (cm)", "Distance To Cover (cm)","Log Diameter (cm)", "Distance To Water (cm)", "Water Depth (cm)", "Percent Water (%)", "Percent Woody Debris (%)", "Percent Floating Vegetation (%)", "Percent Bare Soil (%)", "Mean Temperature (C)", "Eigenvalue", "Explained Variance")

loadings<-cbind(`Variable (units)`,loadings)
names(loadings)<-c("Variable (units)", "Description", "PC-1", "PC-2", "PC-3")
row.names(loadings)<-NULL

kable(loadings) %>%
  kable_styling("striped")%>%
  pack_rows("PCA Statistics", 17,18)
  
```

**Table 2:** Results of the multivariate statistal tests PERMANOVA/PERMDISP based on the Euclidean distance comparing cottonmouth habitat selected and general habitat available to the cottonmouth within the study area.
```{r PERMANOVA/PERMDISP Table, fig.height=4, fig.width=8, message=FALSE, warning=FALSE}
# #PERMANOVA
# Perm1<-adonis2(s1[,c(2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,22)]~ Type, data = s1, method='eu')
# Perm1<-cbind(Perm1$aov.tab[1,4], Perm1$aov.tab[1,6])
# 
# #PERMDISP
# dist<-dist(scale(s1[,c(2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,22)]))
# bd<-anova(betadisper(dist,s1$Type))
# beta<-betadisper(dist,s1$Type)
# bd1<-cbind(bd$`F value`, bd$`Pr(>F)`)
# bd1<-data.frame(bd1)
# bd1<-bd1[-2,]
# PermDisp<- cbind(Perm1, bd1)
# names(PermDisp)<-c("Pseudo-F", "P", "F", "P ")
# row.names(PermDisp)<-c("Proportions (Euclidean)")
# 
# gt(PermDisp, rownames_to_stub = TRUE)%>%
#   fmt_number(columns = c(1,3),
#              decimals = 2,
#              suffixing = T)%>%
#     fmt_scientific(columns = c(2,4),
#              decimals = 2)%>%
#   tab_spanner(
#     label = "PERMANOVA", columns = 1:2)%>%
#   tab_spanner(
#     label = "PERMDISP", columns = 3:4)%>%
#   tab_source_note(
#     source_note = "N(permutations) = 999")%>%
#   tab_source_note(
#     source_note = "df = 1")
```

```{r PERMANOVA/PERMDISP Table GPT version}
vars <- c(2:16, 22)

# distance matrix
d <- dist(scale(s1[, vars]), method = "euclidean")

# PERMANOVA
perm <- adonis2(d ~ Type, data = s1)

Perm1 <- data.frame(
  `Pseudo-F` = perm$F[1],
  P = perm$`Pr(>F)`[1],
  check.names = FALSE
)

# PERMDISP
beta <- betadisper(d, s1$Type)
bd <- anova(beta)

bd1 <- data.frame(
  F = bd$`F value`[1],
  `P ` = bd$`Pr(>F)`[1],
  check.names = FALSE
)

# Combine
PermDisp <- cbind(Perm1, bd1)

row.names(PermDisp) <- "Proportions (Euclidean)"

PermDisp

gt(PermDisp, rownames_to_stub = TRUE)%>%
  fmt_number(columns = c(1,3),
             decimals = 2,
             suffixing = T)%>%
    fmt_scientific(columns = c(2,4),
             decimals = 2)%>%
  tab_spanner(
    label = "PERMANOVA", columns = 1:2)%>%
  tab_spanner(
    label = "PERMDISP", columns = 3:4)%>%
  tab_source_note(
    source_note = "N(permutations) = 999")%>%
  tab_source_note(
    source_note = "df = 1")

```



Below, is the betadisper summary table that shows the average distance to the median of the multivariate space for each cottonmouth and available habitat sites recorded using the PERMDISP statistics.
```{r Betadisper, message=FALSE, warning=FALSE}
beta
```

## Habitat Selection: Intra-specific Differences 

**Table 3:**Mean component scores of PCA on intra-specific habitat selection within cottonmouths with ANOVAs used to compare conspecific groups.

```{r ANOVA Table, message=FALSE, warning=FALSE}
pcomp<-pcomp[,2:4]
names(pcomp)<- c( "PC-1","PC-2", "PC-3")
row.names(pcomp)<-c("Female", "Juvenile", "Male", "YOY")
pcomp[,1:3]<-round(pcomp[,1:3], digits = 3)

C1<-aov(Comp.1~Group, group)
C2<-aov(Comp.2~Group, group)
C3<-aov(Comp.3~Group, group)

F.value<-c(anova(C1)$"F value"[1], anova(C2)$"F value"[1], anova(C3)$"F value"[1])
p.value<-c(anova(C1)$"Pr(>F)"[1],anova(C2)$"Pr(>F)"[1],anova(C3)$"Pr(>F)"[1])

stats<-data.frame(t(cbind(F.value, p.value)))
names(stats)<-c( "PC-1","PC-2", "PC-3")
pcomp<-rbind(pcomp,stats)

row.names(pcomp)<-c("Female", "Juvenile", "Male", "YOY", "F", "P-value")
pcomp<-round(pcomp, digits = 3)

gt(pcomp, rownames_to_stub = TRUE)%>%
  tab_stubhead(label = "Cottonmouth conspecific")%>%
  tab_row_group( group = "",
    rows = 1:4
  )%>%
  tab_row_group(
    group = "ANOVA",rows = c(5,6))
```

Below are the summary tables for the Tukey HSD tests for each principal component.
```{r Tukey HSD, message=FALSE, warning=FALSE}
#PC-1
#ANOVA
summary(C1)
#Tukey
TukeyHSD(C1)
#PC-2
#ANOVA
summary(C2)
#Tukey 2
TukeyHSD(C2)

#PC-3
#ANOVA
summary(C3)

#Tukey 3
TukeyHSD(C3)
```

```{r BoxPlot, fig.height=5, fig.width=10, message=FALSE, warning=FALSE}
PC1<-ggplot(group, aes(x = Group, y = group$Comp.1, color = Group))+
  scale_color_manual(values = c("#F8766D","#7CAE00","#00BFC4", "#C77CFF"))+
  geom_boxplot()+
 ylab(expression(atop("aquatic site" %<->% "terrestrial sites", paste("PC-1 (22.6%)" ))))+
  xlab("")+
  geom_text(aes(4, 2.5, label = paste("P= 0.233")), color = "black", size=4)+
  theme_classic()+
  theme(legend.position="none", legend.background = element_rect(fill="White",size=0.5))

PC2<-ggplot(group, aes(x = Group, y = group$Comp.2, color = Group))+
  scale_color_manual(values = c("#F8766D","#7CAE00","#00BFC4", "#C77CFF"))+
  geom_boxplot()+
  ylab(expression(atop("low cover" %<->% "high cover", paste("PC-2 (11.8%)" ))))+
  xlab("")+
  geom_text(aes(4, 5, label = paste("P= 0.016*")), color = "black", size=4)+
  theme_classic()+
  theme(legend.position="none", legend.background = element_rect(fill="White",size=0.5))

PC3<-ggplot(group, aes(x = Group, y = group$Comp.3, color = Group))+
  scale_color_manual(values = c("#F8766D","#7CAE00","#00BFC4", "#C77CFF"))+
  geom_boxplot()+
  ylab(expression(atop("high vegetation" %<->% "low vegetation", paste("PC-3 (9.2%)" ))))+
  xlab("")+
  geom_text(aes(4, 5, label = paste("P= 0.358")), color = "black", size=4)+
  theme_classic()+
  theme(legend.position="none", legend.background = element_rect(fill="White",size=0.5))

plot_grid(PC1, PC2, PC3, labels=c("A)", "B)", "C)"), ncol = 3, nrow = 1)
```

**Figure:** Boxplots comparing the Principal component scores of each conspecific group for each principal component axes retained from the PCA. Significant difference between conspecific groups is represented with a p-value from an ANOVA test. 

```{r Linear Regression, fig.height= 8, fig.width=22, message=FALSE, warning=FALSE}
#Linear Regression
mod<-lm(SVL~Comp.1,group)
mod2<-lm(SVL~Comp.2,group)
mod3<-lm(SVL~Comp.3,group)

LR1<-ggplot(group, aes(x= SVL, y= Comp.1))+
  geom_point(size = 2)+
  xlim(209, 1061)+
  geom_smooth(method = "lm", color= "black")+
  geom_hline(yintercept = 0, lty = 2) +
  geom_vline(xintercept = 600, lty = 2)+
  theme_bw(base_size = 24)+
  #geom_text(aes(900, 6, label = paste("R-squared = 0.034")), color="black")+
  geom_text(aes(900, 3, label = paste("P = 0.02")), color = "black", size=10, family = "Times New Roman")+
  #geom_text(aes(900, 3, label = paste("P = 0.023")), color = "black", size=10, family = "Times New Roman")+
  xlab("Snout-Vent Length (mm)")+
  ylab(expression(atop("aquatic site" %<->% "terrestrial sites", paste("PC-1 (22.6%)" ))))+
  ylim(-3,3)+
  guides(shape = "none")+
  theme(text = element_text(family = "Times New Roman")) +
  theme(panel.grid = element_blank(), panel.border = element_rect(fill= "transparent"),
    axis.title.x = element_text(size = 28),
    axis.title.y = element_text(size = 28),
    axis.text.x = element_text(size = 22),
    axis.text.y = element_text(size = 22)
  )

LR2<-ggplot(group, aes(x= SVL, y= Comp.2))+
  geom_point(size = 2)+
  xlim(209, 1061)+
  geom_smooth(method = "lm", color= "black")+
  geom_hline(yintercept = 0, lty = 2) +
  geom_vline(xintercept = 600, lty = 2)+
  theme_bw(base_size = 24)+
  #geom_text(aes(900, 6, label = paste("R-squared = 0.045")), color="black")+
  geom_text(aes(900, 3, label = paste("P = 0.008")), color = "black", size = 10, family = "Times New Roman")+
  xlab("Snout-Vent Length (mm)")+
  ylab(expression(atop("low cover" %<->% "high cover", paste("PC-2 (11.8%)" ))))+
  ylim(-3,3)+
  guides(shape = "none")+
  theme(text = element_text(family = "Times New Roman")) +
  theme(panel.grid = element_blank(), panel.border = element_rect(fill= "transparent"),
    axis.title.x = element_text(size = 28),
    axis.title.y = element_text(size = 28),
    axis.text.x = element_text(size = 22),
    axis.text.y = element_text(size = 22)
  )

LR3<-ggplot(group, aes(x= SVL, y= Comp.3))+
  geom_point(size = 2)+
  xlim(209, 1061)+
  geom_hline(yintercept = 0, lty = 2) +
  geom_vline(xintercept = 600, lty = 2)+
  theme_bw(base_size = 24)+
  #geom_text(aes(900, 6, label = paste("R-squared = 0.045")), color="black")+
  geom_text(aes(900, 3, label = paste("P = 0.49")), color = "black", size = 10, family = "Times New Roman")+
  xlab("Snout-Vent Length (mm)")+
  ylab(expression(atop("high vegetation" %<->% "low vegetation", paste("PC-3 (9.2%)" ))))+
  ylim(-3,3)+
  guides(shape = "none")+
  theme(text = element_text(family = "Times New Roman")) +
  theme(panel.grid = element_blank(), panel.border = element_rect(fill= "transparent"),
    axis.title.x = element_text(size = 28),
    axis.title.y = element_text(size = 28),
    axis.text.x = element_text(size = 22),
    axis.text.y = element_text(size = 22)
  )

#combine using patchwork
(LR1 + LR2 + LR3) +
  plot_annotation(tag_levels = 'A')

ggsave("HCB_figs/fig4_SVL_PCA_regressions.jpg")
```

**Figure 2:** Linear regression of Principal Component scores for the comparison of habitat selection among the intra-specific cottonmouth groups(Males, Females, Juveniles, and Neonates) and their snout-vent lengths.

Below are the summary tables for the linear regression analysis comparing body size to each principal component.
```{r Linear Summary, message=FALSE, warning=FALSE}
summary(mod)
summary(mod2)
summary(mod3)
```

## Habitat Selection: Modeling Habitat Preference 

```{r Feature Selection by MIR, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
#Fix characters to factors
s$Type <- as.factor(s$Type)
s1$Type <- as.factor(s1$Type)

rf<-s1[,c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,22)]

set.seed(111)
#Model Improvement Ratio by Cutler et al. 2010
#Train & Test
set.seed(123)
ind<-sample(2,nrow(rf),replace = T, prob = c(0.75,0.25))
train<-rf[ind==1,]
test<-rf[ind==2,]

#Random Forest
RF<-randomForest(Type~.,train, ntree= 6000, importance= T)
RF

#OOB = 16.12; class = 0.151/0.1727

# MIR > 0
A<-data.frame(randomForest::importance(RF))
A
A<-data.frame(A[,2])
sum<-sum(A)
A<-A/sum
A
#Remove 8 Cover Height

#MIR > 0.02
rf<-s[,c(1,2,3,4,5,6,7,9,10,11,12,13,14,15,16,19)]

#Train & Test
set.seed(123)
ind<-sample(2,nrow(rf),replace = T, prob = c(0.75,0.25))
train<-rf[ind==1,]
test<-rf[ind==2,]

#Random Forest
RF<-randomForest(Type~.,train, ntree= 6000, importance= T)
RF

#OOB = 16.94, class= 0.151/0.190
A<-data.frame(randomForest::importance(RF))
A
A<-data.frame(A[,2])
sum<-sum(A)
A<-A/sum
A
# Remove Canopy Cover 1

#MIR > 0.04

rf<-s[,c(1,3,4,5,6,7,9,10,11,12,13,14,15,16,19)]

set.seed(123)
ind<-sample(2,nrow(rf),replace = T, prob = c(0.75,0.25))
train<-rf[ind==1,]
test<-rf[ind==2,]

#Random Forest
RF<-randomForest(Type~.,train, ntree= 6000, importance= T)
RF

#OOB = 17.36, class= 0.151/0.20
A<-data.frame(randomForest::importance(RF))
A
A<-data.frame(A[,2])
sum<-sum(A)
A<-A/sum
A

#Remove 5 WSH, 6 dist, 7 log diam, 12 floating veg

#MIR > 0.06
rf<-s[,c(1,3,4,5,6,11,12,13,14,16,19)]

set.seed(123)
ind<-sample(2,nrow(rf),replace = T, prob = c(0.75,0.25))
train<-rf[ind==1,]
test<-rf[ind==2,]

#Random Forest
RF<-randomForest(Type~.,train, ntree= 6000, importance= T)
RF

#OOB = 18.18, class= 0.189/0.172
A<-data.frame(randomForest::importance(RF))
A
A<-data.frame(A[,2])
sum<-sum(A)
A<-A/sum
A
# Remove Veg, Depth, & Woody Debris

#MIR > 0.08
rf<-s[,c(1,3,5,6,11,13,16,19)]

set.seed(123)
ind<-sample(2,nrow(rf),replace = T, prob = c(0.75,0.25))
train<-rf[ind==1,]
test<-rf[ind==2,]

#Random Forest
RF<-randomForest(Type~.,train, ntree= 6000, importance= T)
RF

#OOB = 21.07, class= 0.181/0.245
A <-data.frame(randomForest::importance(RF))
A
A<-data.frame(A[,2])
sum<-sum(A)
A<-A/sum
A

#Removed Veg, and Fallen Log Cover

#MIR = 0.1
rf<-s[,c(1,3,11,13,16,19)]

set.seed(123)
ind<-sample(2,nrow(rf),replace = T, prob = c(0.75,0.25))
train<-rf[ind==1,]
test<-rf[ind==2,]

#Random Forest
RF<-randomForest(Type~.,train, ntree= 6000, importance= T)
RF

#OOB = 22.31, class= 0.242/0.20
A<-data.frame(randomForest::importance(RF))
A
A<-data.frame(A[,2])
sum<-sum(A)
A<-A/sum
A
```

```{r data, message=FALSE, warning=FALSE}
#Went with the Full Model
rf<-s1[,c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,19)]
```


```{r 10-fold cross validation, message=FALSE, warning=FALSE}
#Setting up Cross Validation Training and Validation sets
set.seed(123)
ind<-sample(2,nrow(rf),replace = T, prob = c(0.9,0.1))
train<-rf[ind==1,]
test<-rf[ind==2,]
```

```{r 30 iterations, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE, cache=TRUE}

train$Type <- as.factor(train$Type)
test$Type  <- as.factor(test$Type)

set.seed(123)
a1<- randomForest(Type~., data = train, ntree = 6000, importance = T)
a2<- randomForest(Type~., data = train, ntree = 6000, importance = T)
a3<- randomForest(Type~., data = train, ntree = 6000, importance = T)
a4<- randomForest(Type~., data = train, ntree = 6000, importance = T)
a5<- randomForest(Type~., data = train, ntree = 6000, importance = T)
a6<- randomForest(Type~., data = train, ntree = 6000, importance = T)
a7<- randomForest(Type~., data = train, ntree = 6000, importance = T)
a8<- randomForest(Type~., data = train, ntree = 6000, importance = T)
a9<- randomForest(Type~., data = train, ntree = 6000, importance = T)
a10<- randomForest(Type~., data = train, ntree = 6000, importance = T)
b1<- randomForest(Type~., data = train, ntree = 6000, importance = T)
b2<- randomForest(Type~., data = train, ntree = 6000, importance = T)
b3<- randomForest(Type~., data = train, ntree = 6000, importance = T)
b4<- randomForest(Type~., data = train, ntree = 6000, importance = T)
b5<- randomForest(Type~., data = train, ntree = 6000, importance = T)
b6<- randomForest(Type~., data = train, ntree = 6000, importance = T)
b7<- randomForest(Type~., data = train, ntree = 6000, importance = T)
b8<- randomForest(Type~., data = train, ntree = 6000, importance = T)
b9<- randomForest(Type~., data = train, ntree = 6000, importance = T)
b10<- randomForest(Type~., data = train, ntree = 6000, importance = T)
c1<- randomForest(Type~., data = train, ntree = 6000, importance = T)
c2<- randomForest(Type~., data = train, ntree = 6000, importance = T)
c3<- randomForest(Type~., data = train, ntree = 6000, importance = T)
c4<- randomForest(Type~., data = train, ntree = 6000, importance = T)
c5<- randomForest(Type~., data = train, ntree = 6000, importance = T)
c6<- randomForest(Type~., data = train, ntree = 6000, importance = T)
c7<- randomForest(Type~., data = train, ntree = 6000, importance = T)
c8<- randomForest(Type~., data = train, ntree = 6000, importance = T)
c9<- randomForest(Type~., data = train, ntree = 6000, importance = T)
c10<- randomForest(Type~., data = train, ntree = 6000, importance = T)
```

```{r Error Rate Plot, eval=FALSE, include=FALSE}
#Error rate of Random Forest
plot(a1)
```

```{r Out-of-Bag, message=FALSE, warning=FALSE}
OOB1<-c(a1$err.rate[6000,1],a2$err.rate[6000,1],a3$err.rate[6000,1],a4$err.rate[6000,1],a5$err.rate[6000,1],a6$err.rate[6000,1],
       a7$err.rate[6000,1],a8$err.rate[6000,1],a9$err.rate[6000,1],a10$err.rate[6000,1],b1$err.rate[6000,1],b2$err.rate[6000,1],
       b3$err.rate[6000,1],b4$err.rate[6000,1],b5$err.rate[6000,1],b6$err.rate[6000,1],b7$err.rate[6000,1],b8$err.rate[6000,1],
       b9$err.rate[6000,1],b10$err.rate[6000,1],c1$err.rate[6000,1],c2$err.rate[6000,1],c3$err.rate[6000,1],c4$err.rate[6000,1],
       c5$err.rate[6000,1],c6$err.rate[6000,1],c7$err.rate[6000,1],c8$err.rate[6000,1],c9$err.rate[6000,1],c10$err.rate[6000,1])
OOB<-c(1-min(OOB1),1-mean(OOB1), 1-max(OOB1))
```

```{r Cross-Validation, message=FALSE, warning=FALSE}
files<-list(a1,a2,a3,a4,a5,a6,a7,a8,a9,a10,b1,b2,b3,b4,b5,b6,b7,b8,b9,b10,c1,c2,c3,c4,c5,c6,c7,c8,c9,c10)

cv<-(function(x){
p<-predict(x, test)
cm<-caret::confusionMatrix(p, positive = "SNAKE",test$Type)

CVTABLE<-t(data.frame(cm$overall))
CV2TABLE<-t(data.frame(cm$byClass))

cv_table<-cbind(CVTABLE,CV2TABLE)
cv_table<-cv_table[,c(1,2,8,9)]
cv_table
})

cv_table<-t(data.frame(lapply(files, cv)))

cv_max<-c(max(cv_table[,1]), max(cv_table[,2]), max(cv_table[,3]), max(cv_table[,4]))
cv_mean<-c(mean(cv_table[,1]), mean(cv_table[,2]), mean(cv_table[,3]), mean(cv_table[,4]))
cv_min<-c(min(cv_table[,1]), min(cv_table[,2]), min(cv_table[,3]), min(cv_table[,4]))

cv_table<-data.frame(rbind(cv_max,cv_mean, cv_min))

RandomForest_Model<-c("Maximum", "Mean", "Minimum")
```

```{r Calculating AUC, message=FALSE, warning=FALSE}

AUC<- function(z){
pred<-predict(z, test, type = "prob")
pred<-pred[,2]
pred<-prediction(pred, test$Type)

#AUC (Area Under The Curve)
auc<-performance(pred, "auc")
auc<-unlist(slot(auc, "y.values"))
  
}

AUC<-t(data.frame(lapply(files, AUC)))

AUC<-c(max(AUC), mean(AUC), min(AUC))
```

**Table 4:** Performance metrics for the 30-iteration random forest model on cottonmouth habitat selection compared to general habitat available within the study area.

```{r Random Forest Table, message=FALSE, warning=FALSE}
models<-cbind(cv_table, OOB, AUC)
models<-round(models, digits = 3)
models<-cbind(RandomForest_Model,models)
gt(models)%>%
  cols_label(
    RandomForest_Model = "Random Forest Models",
    X1 = "Accuracy",
    X2 = "Kappa",
    X3 = "Sensitivity",
    X4 = "Specificity")%>%
  tab_source_note(
    source_note = "N(trees) = 6000")%>%
  tab_source_note(
    source_note = "OOB = Out of Bag Accuracy")%>%
    tab_source_note(
    source_note = "AUC = Area Under the Curve")
```

Below are the summary tables for the maximum and mean random forest models of the 30-iteration models.
```{r Maximum RF Model, message=FALSE, warning=FALSE}
a7 # Maximum Accuracy
a1 #Mean Accuracy
```

```{r Variable Importance, fig.height=10, fig.width=15, message=FALSE, warning=FALSE}
MDA<-randomForest::importance(a7, class = "SNAKE")
MDA<-data.frame(MDA[,c(1,2)])
gen<-sum(MDA[,1])
sna<-sum(MDA[,2])
#Calculating relative importance of MDA
MDA[,1]<-(MDA[,1]/gen)*100
MDA[,2]<-(MDA[,2]/gen)*100

#Plotting
Habitat_Variables <- c("Canopy Closure", "Rock Cover", "Leaf Cover", "Vegetative Cover","Fallen Log Cover","Woody Stem Height", "Cover Height", "Distance To Cover","Log Diameter","Distance To Water", "Water Depth", "Percent Water", "Percent Woody Debris", "Percent Floating Vegetation", "Percent Bare Soil", "Mean Temperature")

D <-cbind(Habitat_Variables,MDA)
row.names(D)<-NULL

ggplot(D, aes(x = reorder(Habitat_Variables, D$SNAKE), y = D$SNAKE))+
  geom_col(aes(fill=D$SNAKE), color = "black", width = 0.9)+
  geom_text(label=round(D$SNAKE,1), nudge_y = 0.5)+
    ylab("Relative Importance")+
  xlab(expression(atop("Environmental Variables", paste(""))))+
  coord_flip()+
  theme_classic()+
  theme(text = element_text(family = "Times New Roman")) +
  theme(legend.position = "none",
    axis.title.x = element_text(size = 24),
    axis.title.y = element_text(size = 24),
    axis.text.x = element_text(size = 20),
    axis.text.y = element_text(size = 20, color = "black")
  ) +
  scale_fill_gradient2(low="white", high="black")

ggsave("HCB_figs/fig5_variable_importance.jpg")
```



```{r Partial Dependence Plots Set up, message=FALSE, warning=FALSE}
#RF a7
plotx<-partial(a7, pred.var = "Dist..To.Water", pdp = T, center = T, plot = T, plot.engine = "ggplot2", type = "classification", which.class = "SNAKE", train = train, rug = T)

ploty<-partial(a7, pred.var = "X.Bare.soil", pdp = T, center = T, plot = T, plot.engine = "ggplot2", type = "classification", which.class = "SNAKE", train = train, rug = T)

plotz<-partial(a7, pred.var = "Rock.Cover", pdp = T, center = T, plot = T, plot.engine = "ggplot2", type = "classification", which.class = "SNAKE", train = train, rug = T)

plotz1<-partial(a7, pred.var = "Depth", pdp = T, center = T, plot = T, plot.engine = "ggplot2", type = "classification", which.class = "SNAKE", train = train, rug = T)

plotz2<-partial(a7, pred.var = "X.Water", pdp = T, center = T, plot = T, plot.engine = "ggplot2", type = "classification", which.class = "SNAKE", train = train, rug = T)

plotz3<-partial(a7, pred.var = "Dist..To.Cover", pdp = T, center = T, plot = T, plot.engine = "ggplot2", type = "classification", which.class = "SNAKE", train = train, rug = T)
```

```{r Partial Dependence Plots, message=FALSE, warning=FALSE}

plot1<-ploty+ geom_line(size = 1)+
  xlab("Bare Soil Cover (%)")+
  ylab("Relative Probability")+
  xlim(0,105)+
  theme_bw()+
  theme(text = element_text(family = "Times New Roman")) +
  theme(axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(size = 10, color = "black"),
    axis.text.y = element_text(size = 10, color = "black")
  )

plot2<-plotx+geom_line(size = 1)+
  xlab("Distance to Water (cm)")+
  ylab("")+
  ylim(-1.05,0.35)+
  theme_bw()+
  theme(text = element_text(family = "Times New Roman")) +
  theme(axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(size = 10, color = "black"),
    axis.text.y = element_text(size = 10, color = "black")
  )

plot3<-plotz+ geom_line(size = 1)+
  xlab("Rock Cover (%)")+
  ylab("")+
  theme_bw()+
  theme(text = element_text(family = "Times New Roman")) +
  theme(axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(size = 10, color = "black"),
    axis.text.y = element_text(size = 10, color = "black")
  )

plot4<-plotz1+ geom_line(size = 1)+
  xlab("Water Depth (cm)")+
  ylab("Relative Probability")+
  xlim(0,105)+
  ylim(-0.25,0.4)+
  theme_bw()+
  theme(text = element_text(family = "Times New Roman")) +
  theme(axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(size = 10, color = "black"),
    axis.text.y = element_text(size = 10, color = "black")
  )

plot5<-plotz2+ geom_line(size = 1)+
  xlab("Water Cover (%)")+
  ylab("")+
  xlim(0,105)+
  ylim(-0.80, 0.25)+
  theme_bw()+
  theme(text = element_text(family = "Times New Roman")) +
  theme(axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(size = 10, color = "black"),
    axis.text.y = element_text(size = 10, color = "black")
  )

plot6<-plotz3+ geom_line(size = 1)+
  xlab("Distance to Cover (cm)")+
  ylab("")+
  ylim(-0.65,-0.2)+
  theme_bw()+
  theme(text = element_text(family = "Times New Roman")) +
  theme(axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text.x = element_text(size = 10, color = "black"),
    axis.text.y = element_text(size = 10, color = "black")
  )

#combine plots using patchwork
(plot1 + plot2 + plot3) / (plot4 + plot5 + plot6) +
  plot_annotation(tag_levels = 'A')

ggsave("HCB_figs/fig6_partial_dependence_plots.jpg")

```


**Figure 3:** The Relative Importance (based on the Mean Decrease Accuracy) on the environmental variables from the minimum error rate random forest model on Cottonmouth (Agkistrodon piscivorus) habitat selection in Middle TN, USA. Cover variables were measured in percentage of occurance within a 1-m^2 quadrat, and all other metrics were measured in cm.

```{r Partial Dependance Plots, fig.height=8, fig.width=15, message=FALSE, warning=FALSE}
#plot_grid(plot1, plot2,plot3,plot4,plot5,plot6, labels=c("A ", "B ", "C ","D ", "E ", "F "), ncol = 3, nrow = 2)
```

**Figure 4:** Partial dependance plots of the four most constributing variables (From mean decrease accuracy) with relative probability of cottonmouth habitat selection of a site within Middle Tennessee, USA. Distance to water (A), bare soil cover (B), rock cover (C), and water depth (D) were measured within and above a 1-m^2 quadrat from were a cottonmouth was located during visual encounter. Dashes on the x-axis indicate the deciles of the cottonmouth occurence for each environmental variables.


